---
title: "Life Expectancy: An Analysis and Prediction Based on Key Influencing Factors"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_caption: true
  html_document:
    df_print: paged
---

** Import Library**
```{r,echo=FALSE}
library(ggplot2)
library(dplyr)
```

** Read Data File**
```{r,echo=FALSE}
data=read.csv("./data/Life-Expectancy-Data-Updated.csv")

data$Year[which(data$Year==2000)]="Y2000"
data$Year[which(data$Year==2001)]="Y2001"
data$Year[which(data$Year==2002)]="Y2002"
data$Year[which(data$Year==2003)]="Y2003"
data$Year[which(data$Year==2004)]="Y2004"
data$Year[which(data$Year==2005)]="Y2005"
data$Year[which(data$Year==2006)]="Y2006"
data$Year[which(data$Year==2007)]="Y2007"
data$Year[which(data$Year==2008)]="Y2008"
data$Year[which(data$Year==2009)]="Y2009"
data$Year[which(data$Year==2010)]="Y2010"
data$Year[which(data$Year==2011)]="Y2011"
data$Year[which(data$Year==2012)]="Y2012"
data$Year[which(data$Year==2013)]="Y2013"
data$Year[which(data$Year==2014)]="Y2014"
data$Year[which(data$Year==2015)]="Y2015"

data$Economy_status_Developed[which(data$Economy_status_Developed==0)]="No"
data$Economy_status_Developed[which(data$Economy_status_Developed==1)]="Yes"

head(data,10)
options(max.print = 9999)
```

```{r,echo=FALSE}
# Check summary statistics for the key variables
summary(data[, c("Life_expectancy", "Infant_deaths", "Adult_mortality", "Economy_status_Developed")])

```

**Data Distribution:**



```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = 'GDP_per_capita')) +
  geom_histogram(binwidth = 2500, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of GDP Per Capita"), x = "GDP Per Capita in USD", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```



```{r}

# Create the plot
plot <- ggplot(data, aes_string(x = 'Infant_deaths')) +
  geom_histogram(binwidth = 3, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Infant Deaths"), x = "Infant Deaths per 1000 population", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```





```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Adult_mortality")) +
  geom_histogram(binwidth = 14, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Adult Mortality"), x = "Adult Mortality per 1000 population", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```





```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Alcohol_consumption")) +
  geom_histogram(binwidth = 0.4, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of", "Alcohol Consumption"), x = "Alcohol Consumption in liters per capita with 15+ years old", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```





```{r}

# Create the plot
plot <- ggplot(data, aes_string(x = "BMI")) +
  geom_histogram(binwidth = 0.3, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of BMI"), x = "BMI", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```





```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Life_expectancy")) +
  geom_histogram(binwidth = 1.2, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Life Expectancy"), x = "Life Expectancy in Years", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```


```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = 'Hepatitis_B')) +
  geom_histogram(binwidth = 2.4, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Hepatitis B"), x = "Hepatitis B Vaccination Rate among 1-year-olds", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```



```{r}
  # List of numeric columns for histograms
columns <- c(
   "Measles"
)

  # Create the plot
  plot <- ggplot(data, aes_string(x = "Measles")) +
    geom_histogram(binwidth = 2.5, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of Measles"), x = "Measles Vaccination Rate among 1-year-olds", y = "Frequency") +
    theme_minimal()
  
  # Print the plot
  print(plot)

```



```{r}
# Create the plot
  plot <- ggplot(data, aes_string(x = "Polio")) +
    geom_histogram(binwidth = 2.5, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of Polio"), x = "Polio Vaccination Rate among 1-year-olds", y = "Frequency") +
    theme_minimal()
  # Print the plot
  print(plot)
```


```{r}

# Create the plot
plot <- ggplot(data, aes_string(x = "Diphtheria")) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Diphtheria"), x = "Diphtheria Vaccination Rate among 1-year-olds", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```




```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Population_mln")) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Population"), x = "Population in millions", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```



```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Thinness_ten_nineteen_years")) +
  geom_histogram(binwidth = 0.6, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Thinness Ten Nineteen Years"), x = "Thinness Ten Nineteen Years where BMI < -2 standard deviations below the median", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```




```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Thinness_five_nine_years")) +
  geom_histogram(binwidth = 0.7, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Thinness Ten Nineteen Years"), x = "Thinness Five to Nine Years where BMI < -2 standard deviations below the median", y = "Frequency") +
theme_minimal()

# Print the plot
print(plot)
```




```{r}
# Create the plot
plot <- ggplot(data, aes_string(x = "Schooling")) +
  geom_histogram(binwidth = 0.25, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of Schooling"), x = "Average years that people aged 25+ spent in formal education", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```


```{r}

# Create the plot
plot <- ggplot(data, aes_string(x = "Incidents_HIV")) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = paste("Histogram of HIV Incidents"), x = "Incidents of HIV per 1000 population aged 15-49", y = "Frequency") +
  theme_minimal()

# Print the plot
print(plot)

```

**Relatioship between life Expectancy and Infant Deaths**

```{r,echo=FALSE}
# Scatter plot to visualize relationship between life expectancy and infant deaths
ggplot(data, aes(x = Infant_deaths, y = Life_expectancy)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Life Expectancy vs. Infant Deaths",
       x = "Infant Deaths per 1000 Population",
       y = "Life Expectancy (years)") +
  theme_minimal()

```

**Relatioship between life Expectancy and Adult Mortality**

```{r,echo=FALSE}
# Scatter plot to visualize relationship between life expectancy and adult mortality
ggplot(data, aes(x = Adult_mortality, y = Life_expectancy)) +
  geom_point(color = "green", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Life Expectancy vs. Adult Mortality",
       x = "Adult Mortality Rate (per 1000)",
       y = "Life Expectancy (years)") +
  theme_minimal()

```

**Relatioship between life Expectancy and economy status (Developed vs. Developing**

```{r,echo=FALSE}
# Box plot to visualize relationship between life expectancy and economy status (Developed vs. Developing)
ggplot(data, aes(x = Economy_status_Developed, y = Life_expectancy, fill = factor(Economy_status_Developed))) +
  geom_boxplot() +
  scale_x_discrete(labels = c("Not Developed", "Developed")) +
  labs(title = "Life Expectancy by Economy Status",
       x = "Economy Status",
       y = "Life Expectancy (years)") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "lightgreen"))

```


```{r,echo=FALSE}
# Full model
model = lm(Life_expectancy ~ Region + Year + Infant_deaths + Under_five_deaths + Adult_mortality + Hepatitis_B + Alcohol_consumption + Measles + BMI + Polio + Diphtheria + Incidents_HIV + GDP_per_capita + Population_mln + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed, data=data)
library(mctest)
imcdiag(model, method="VIF")
```

```{r,echo=FALSE}
# 
library(GGally)
ggpairs(data, columns = c("Life_expectancy", "Infant_deaths", "Under_five_deaths",  "Polio", "Diphtheria"))
```

```{r,echo=FALSE}
# Model after getting rid of multi-collinearlity terms
model = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + Hepatitis_B + Alcohol_consumption + Measles + BMI + Polio + Incidents_HIV + GDP_per_capita + Population_mln + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)
imcdiag(model, method="VIF")
```

```{r,echo=FALSE}
# Full f-test
null_model=lm(Life_expectancy~1,data=data)
anova(null_model,model)
```

\label{Figuere 3.2} 
$$Full F - Test (Anova Test)$$

```{r,echo=FALSE}
# Individual t-test
summary(model)
```

```{r,echo=FALSE}
# Model after getting rid of insignificant terms
first_order_model = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)
summary(first_order_model)
```

```{r,echo=FALSE}
# Using step-wise to compare our models
library(olsrr)

step_model = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)

summary(ols_step_both_p(step_model,p_enter=0.1,p_remove=0.3)$model)
```

```{r,echo=FALSE}
# Try interactive models 
int_model = lm(Life_expectancy ~ (Region + Year + Infant_deaths + Adult_mortality + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed)^2,data=data)
summary(int_model)
```

```{r,echo=FALSE}
# Interactive model after removing insignificant interactive terms
int_model_2 = lm(Life_expectancy ~ Region +
               Year +
               Infant_deaths + 
               Adult_mortality + 
               Hepatitis_B + BMI +
               Polio + 
               Incidents_HIV + 
               GDP_per_capita + 
               Thinness_ten_nineteen_years +
               Thinness_five_nine_years +
               Schooling + 
               Economy_status_Developed + 
               Region:Year +  
               Region:Infant_deaths + 
               Region:Adult_mortality +
               Region:Hepatitis_B + 
               Region:BMI +
               Region:Polio +
               Region:Incidents_HIV + 
               Region:GDP_per_capita + 
               Region:Thinness_ten_nineteen_years  + 
               Region:Thinness_five_nine_years + 
               Region:Schooling + 
               Region:Economy_status_Developed + 
               Year:BMI + 
               Year:Incidents_HIV + 
               Year:Thinness_ten_nineteen_years + 
               Year:Thinness_five_nine_years + 
               Infant_deaths:Adult_mortality + 
               Infant_deaths:Hepatitis_B + 
               Infant_deaths:BMI + 
               Infant_deaths:GDP_per_capita + 
               Adult_mortality:Hepatitis_B + 
               Adult_mortality:BMI + 
               Adult_mortality:Polio + 
               Adult_mortality:Incidents_HIV + 
               Adult_mortality:GDP_per_capita + 
               Adult_mortality:Schooling + 
               Hepatitis_B:Polio + 
               Hepatitis_B:Incidents_HIV + 
               Hepatitis_B:Thinness_five_nine_years + 
               BMI:Thinness_ten_nineteen_years +
               BMI:Thinness_five_nine_years +
               BMI:Schooling + 
               Polio:Incidents_HIV +
               Incidents_HIV:GDP_per_capita +
               Incidents_HIV:Thinness_ten_nineteen_years +
               Incidents_HIV:Thinness_five_nine_years +
               Thinness_ten_nineteen_years:Thinness_five_nine_years , data= data) 
summary(int_model_2)
```

```{r,echo=FALSE}
# again doing the same as above
int_model_3 = lm(Life_expectancy ~ Region +
               Year +
               Infant_deaths + 
               Adult_mortality + 
               Hepatitis_B + BMI +
               Polio + 
               Incidents_HIV + 
               GDP_per_capita + 
               Thinness_ten_nineteen_years +
               Thinness_five_nine_years +
               Schooling + 
               Economy_status_Developed + 
               Region:Year +  
               Region:Infant_deaths + 
               Region:Adult_mortality +
               Region:Hepatitis_B + 
               Region:BMI +
               Region:Polio +
               Region:Incidents_HIV + 
               Region:GDP_per_capita + 
               Region:Thinness_ten_nineteen_years  + 
               Region:Thinness_five_nine_years + 
               Region:Schooling + 
               Region:Economy_status_Developed + 
               Year:BMI + 
               Year:Incidents_HIV + 
               Year:Thinness_ten_nineteen_years + 
               Year:Thinness_five_nine_years + 
               Infant_deaths:Adult_mortality + 
               Infant_deaths:Hepatitis_B + 
               Infant_deaths:BMI + 
               Infant_deaths:GDP_per_capita + 
               Adult_mortality:Hepatitis_B + 
               Adult_mortality:BMI + 
               Adult_mortality:Polio + 
               Adult_mortality:Incidents_HIV + 
               Adult_mortality:GDP_per_capita + 
               Adult_mortality:Schooling + 
               Hepatitis_B:Incidents_HIV + 
               Hepatitis_B:Thinness_five_nine_years + 
               BMI:Thinness_ten_nineteen_years +
               BMI:Thinness_five_nine_years +
               BMI:Schooling + 
               Polio:Incidents_HIV +
               Incidents_HIV:GDP_per_capita +
               Incidents_HIV:Thinness_ten_nineteen_years +
               Incidents_HIV:Thinness_five_nine_years , data= data) 
summary(int_model_3)
```

```{r,echo=FALSE}
library(GGally)
# Finding the variable with highest significance in the model
ggpairs(data, columns = c("Life_expectancy", "Region", "Year", "Infant_deaths", "Adult_mortality", 
"Hepatitis_B", "BMI", "Polio", "Incidents_HIV", "GDP_per_capita", 
"Thinness_ten_nineteen_years", "Thinness_five_nine_years", "Schooling", 
"Economy_status_Developed"),cardinality_threshold = 30)
```

```{r,echo=FALSE}
# Second order model with (Adult_mortality)
sec_order_model_1 = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + I(Adult_mortality^2) + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed, data= data) 
summary(sec_order_model_1)
```

```{r,echo=FALSE}
# Second order model with (Infant_deaths)
sec_order_model_2 = lm(Life_expectancy ~ Region + Year + Infant_deaths + I(Infant_deaths^2) + Adult_mortality + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed, data= data) 
summary(sec_order_model_2)
```

```{r,echo=FALSE}
#All-Possible-Regressions Selection Procedure
library(leaps)
subsets=regsubsets(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed, data= data, nv=13)
subsets2=ols_step_best_subset(step_model)
sm=summary(subsets)
rsquare=c(sm$rsq)
AdjustedR=c(sm$adjr)
cp=c(sm$cp)
AIC=c((subsets2$metrics)$aic)
BIC=c(sm$bic)
cbind(rsquare,AdjustedR,cp,AIC,BIC)
```

```{r}
valid_model = first_order_model
```

**Residual Analysis:**

Residual analysis is a critical step in validating and assessing the quality of a regression model. Residuals, which represent the differences between the observed data points and the corresponding values predicted by the model, provide insights into whether the foundational assumptions of regression analysis are being met. Properly addressing these assumptions is essential to ensure that the model delivers accurate and reliable results.


Key assumptions in regression analysis are given below:

1) Linearity Assumption
2) Independence Assumption
3) Equal Variance Assumption
4) Normality Assumption
5) Outliers

Violations of these assumptions can compromise the model’s reliability. Residual analysis involves using visualizations like residual plots and statistical tests to detect issues.

In analyzing life expectancy, residual analysis is crucial due to the intricate relationships between health, economic, and social factors. Ensuring that regression assumptions are satisfied strengthens the model's credibility and enhances the reliability of its predictions. This process provides valuable insights into the key drivers influencing life expectancy, enabling more informed decision-making and targeted interventions.


**1) Linearity Assumption:**
In regression analysis, the linearity assumption requires that the dependent variable maintains a straight-line relationship with the independent variables. In other words, the predicted values of the dependent variable should change proportionally with the predictors. If the relationship is non-linear, this assumption is violated, potentially resulting in biased estimates and less reliable predictions.

To assess whether this assumption holds, residual plots are frequently utilized. These plots compare the residuals against fitted values or independent variables. When residuals are randomly dispersed around zero without discernible patterns, it suggests the linearity assumption is met. On the other hand, noticeable patterns—such as curves or trends—signal a non-linear relationship that may require adjustments to the model.


```{r,echo=FALSE}


# Load ggplot2
library(ggplot2)

#Plot Fitted v/s Residual
ggplot(valid_model, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0) + 
labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

Based on this we can say that there is some non linear pattern indicate ath relationship between life expectancy and independent variables are not linear, lets add some higher order term.

```{r,echo=FALSE}
library(GGally)

ggpairs(data, columns = c("Life_expectancy", "Region", "Infant_deaths", "Adult_mortality",  "Hepatitis_B", "BMI", "Polio", "Incidents_HIV", "GDP_per_capita", "Thinness_ten_nineteen_years", "Thinness_five_nine_years", "Schooling", "Economy_status_Developed"))

```

Based on this We can say that Adult mortality has higher correlation so that we start with it.


```{r,echo=FALSE}

higher_order_model = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + I(Adult_mortality^2) + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)


summary(higher_order_model)

```
```{r,echo=FALSE}
#Plot Fitted v/s Residual
ggplot(higher_order_model, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0) + 
labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

```

Based on the graph, we can observe that it is becoming more aligned with the zero line. Let us increase the order and check if the graph aligns even more closely with zero.

```{r,echo=FALSE}

higher_order_model_2 = lm(Life_expectancy ~ Region + Year + Infant_deaths + Adult_mortality + I(Adult_mortality^2) + I(Adult_mortality^3) + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)



summary(higher_order_model_2)
```
```{r,echo=FALSE}
#Plot Fitted v/s Residual
ggplot(higher_order_model_2, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0) + 
labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

Based on graph, we can not reduce the non linearity of curve so we can say that there is no further reduction is possible with Adult_mortality term, lets check with other highly correlated term i.e. Infant_deaths.

```{r,echo=FALSE}

higher_order_model_3 = lm(Life_expectancy ~ Region + Year + Infant_deaths + I(Infant_deaths^2) + Adult_mortality + I(Adult_mortality^2) + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)

summary(higher_order_model_3)


```
```{r,echo=FALSE}
#Plot Fitted v/s Residual
ggplot(higher_order_model_3, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0) + 
labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

```

Based on the graph, there appears to be a subtle pattern in the residuals, though it is less pronounced. Lets check for cubic model.

```{r,echo=FALSE}


higher_order_model_4 = lm(Life_expectancy ~ Region + Year + Infant_deaths + I(Infant_deaths^2) + I(Infant_deaths^3) + Adult_mortality + I(Adult_mortality^2) + Hepatitis_B + BMI + Polio + Incidents_HIV + GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed,data=data)

summary(higher_order_model_4)


```

```{r,echo=FALSE}

#Plot Fitted v/s Residual
ggplot(higher_order_model_4, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0) + 
labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

The residuals appear to be randomly scattered, showing no discernible pattern, which suggests that the linearity assumption is satisfied. The $R_{adj}^2$  value of 0.9846 indicates that 98.46% of the variation in life expectancy is explained by this model. Additionally, the Root Mean Squared Error (RMSE) of 1.167 reflects the average deviation of the predicted values from the observed values, demonstrating the model's high accuracy in predicting life expectancy.
```{r,echo=FALSE}
valid_model = higher_order_model_4
```


**2) Independence Assumption:**

The Independence Assumption in regression analysis states that the residuals (errors) in the model should be independent of each other. This means that the error associated with one observation should not provide any information about the error associated with another observation. This typically occurs when the data for both dependent and
independent variables are observed sequentially over a period of time-called time-series data. Since our data does not related to time, space, or group so we can be pretty sure that their measurements are independent.


**3) Equal Variance Assumption:**

The Equal Variance Assumption, also known as Homoscedasticity, is a fundamental assumption in regression analysis. It states that the variance of the residuals (errors) should be constant across all levels of the independent variables. This is crucial for the validity of statistical tests and confidence intervals, as violations can lead to biased estimates of regression coefficients, incorrect standard errors, and unreliable hypothesis tests.

We will check this using Breusch-Pagan Test,

Null Hypothesis $H_{0}$: There is no heteroscedasticity is present, i.e. $\sigma_1^2 = \sigma_2^2 = ... =\sigma_n^2$

Alternative Hypothesis $H_{1}$: heteroscedasticity is present, atleast one $\sigma_i^2$ is different from others i= 1,2,..n.

```{r,echo=FALSE}
library(lmtest)

bptest(valid_model)
```
A significant result, p-value < 2.2e-16 which is less than 0.05 from the Breusch-Pagan test suggests heteroscedasticity, meaning the variance of the residuals is not constant across levels of the independent variables. We reject null hypothesis.

Equal Variance Assumption is violated, a Box-Cox Transformation can stabilize the variance. This method identifies an appropriate transformation for the dependent variable.

```{r,echo=FALSE}

library(MASS)

# Apply Box-Cox transformation
boxcox_model = boxcox(valid_model, lambda = seq(-1, 1))

# Identify the optimal lambda (power transformation)
optimal_lambda = boxcox_model$x[which.max(boxcox_model$y)]
cat("Optimal Lambda for Box-Cox Transformation:", optimal_lambda, "\n")

```

When the Box-Cox transformation identifies an optimal lambda value of 0.59596, it suggests that the dependent variable (e.g., Life_Expectancy) requires a power transformation to stabilize variance and meet the regression assumptions.


```{r,echo=FALSE}
model_transformed = lm(formula = (((Life_expectancy^0.59596)-1)/0.59596) ~ Region + Year + Infant_deaths + 
    I(Infant_deaths^2) + I(Infant_deaths^3) + Adult_mortality + 
    I(Adult_mortality^2) + Hepatitis_B + BMI + Polio + Incidents_HIV + 
    GDP_per_capita + Thinness_ten_nineteen_years + Thinness_five_nine_years + 
    Schooling + Economy_status_Developed, data = data)
  
summary(model_transformed)
```
Let's check homoscedasticity for transformed model using Breusch-Pagan Test,
Null Hypothesis $H_{0}$: There is no heteroscedasticity is present, i.e. $\sigma_1^2 = \sigma_2^2 = ... =\sigma_n^2$

Alternative Hypothesis $H_{1}$: heteroscedasticity is present, atleast one $\sigma_i^2$ is different from others i= 1,2,..n.


```{r,echo=FALSE}
bptest(model_transformed)

```
A significant result, p-value < 2.2e-16 which is less than 0.05 from the Breusch-Pagan test suggests heteroscedasticity, meaning the variance of the residuals is not constant across levels of the independent variables. We reject null hypothesis.

After applying the Box-Cox transformation in an attempt to resolve the issue of heteroscedasticity, the Breusch-Pagan test confirmed that heteroscedasticity still persists in the residuals, indicating that the transformation was not sufficient to address the variance instability.

**4) Normality Assumption:**

The Normality Assumption in regression analysis requires that the residuals of the model are normally distributed. We will verify using a Q-Q-Plot.

```{r,echo=FALSE}
#normal QQ plot
ggplot(valid_model, aes(sample=valid_model$residuals)) +
stat_qq() +
stat_qq_line()

```
Plotted Q-Q plot, indicate that the residuals follow a normal distribution. This suggests that the normality assumption of the regression model is satisfied.

Lets perform, Shapiro-Wilk test

Null Hypothesis $H_{0}$: The sample data are significantly normally distributed
Alternative Hypothesis $H_{1}$: The sample data are not significantly normally distributed

```{r,echo=FALSE}
shapiro.test(residuals(valid_model))
```
The Shapiro-Wilk normality test further supports that the residuals are normally distributed, as the p-value (0.06246) is greater than 0.05. Therefore, we fail to reject the null hypothesis, indicating that the residuals follow a normal distribution.

**5) Outliers:**

Outliers are data points that deviate significantly from the overall pattern of the data. In regression analysis, outliers can have a disproportionate influence on the estimated coefficients, residuals, and overall model fit.

We will be using Cook distance to measure of the overall influence an outlying observation.

```{r,echo=FALSE}

plot(valid_model,pch=18,col="red",which=c(4)) #which =4 only prints the cook distance plot.

```

Based on the graph, there are no outliers identified using Cook's Distance with a threshold of 0.5. This indicates that no observations exceed this threshold and are deemed influential in the context of the model.

--------------------------------------------------------------

After reviewing all residual assumptions and making necessary modifications to the model, we can conclude that the final model is valid and satisfies the key regression assumptions, ensuring its reliability for analysis and prediction.



$\hat{Life_expectancy} = 83.846 + 0.5985 * Region_{Asia} + 2.134 * Region_{Central America and Caribbean} - 0.5379 * Region_{European Union} + 0.07193 * Region_{Middle East} + 0.7183 * Region_{North America} - 0.4468 * Region_{Oceania} + 0.3035 * Region_{Rest of Europe} - 1.885 * Region_{South America} + 0.06593 * Year_{Y2001} + 0.06393 * Year_{Y2002} + 0.004886 * Year_{Y2003} + 0.05015 * Year_{Y2004} + 0.01471 * Year_{Y2005} + 0.04960 * Year_{Y2006} + 0.08880 * Year_{Y2007} + 0.1522 * Year_{Y2008} + 0.2178 * Year_{Y2009} + 0.2926 * Year_{Y2010} + 0.3463 * Year_{Y2011} + 0.3679 * Year_{Y2012} + 0.4794 * Year_{Y2013} + 0.5691 * Year_{Y2014} + 0.5864 * Year_{Y2015} - 0.1189 * InfantDeaths - 0.0003900 * InfantDeaths^2 + 0.000003453 * InfantDeaths^3 - 0.06159 * AdultMortality + 0.00002649 * AdultMortality^2 + 0.01159 * HepatitisB - 0.1046 * BMI + 0.01180 * Polio - 0.03977 * IncidentsHIV + 0.000007030 * GDP_per_capita - 0.04017 * ThinnessTenNineteenYears + 0.04967 * ThinnessFiveNineYears + 0.1097 * Schooling + 2.084 * EconomyStatusDeveloped_{Yes}$


**Prediction:**

```{r}
#Based on Economy Status
# For Canada 
newdata = data.frame(Region="North America", Year="Y2014", Infant_deaths=5.3, Under_five_deaths=6.3, Adult_mortality=80.9805, Alcohol_consumption=7.6, Hepatitis_B=42, Measles=87,  BMI=26.5, Polio=94, Diphtheria=80, Incidents_HIV=0.1, GDP_per_capita=43635, Population_mln=34.44, Thinness_ten_nineteen_years=0.5, Thinness_five_nine_years=0.5, Schooling=13, Economy_status_Developed="Yes", Economy_status_Developing="No")
predict(valid_model, newdata, type="response")

# For Mexico
newdata = data.frame(Region="North America", Year="Y2014", Infant_deaths=19.2, Under_five_deaths=21.5, Adult_mortality=123, Alcohol_consumption=5.21, Hepatitis_B=98, Measles=87,  BMI=27.1, Polio=99, Diphtheria=91, Incidents_HIV=0.15, GDP_per_capita=8703, Population_mln=117.27, Thinness_ten_nineteen_years=1.6, Thinness_five_nine_years=1,7, Schooling=8.7, Economy_status_Developed="No", Economy_status_Developing="Yes")
predict(valid_model, newdata, type="response")
```



This analysis highlights the significant role that Economy Status plays in determining life expectancy. The model predicts that Canada, with its developed economy and strong healthcare system, has a significantly higher life expectancy than Mexico, a developing country with greater healthcare challenges and lower economic indicators. These predictions reinforce the importance of economic development, healthcare access, and education in shaping a country’s overall life-expectancy.
